Co-evolution of lexical meaning \& pragmatic use

### 2017-03-23 paper/paper-v3.tex
paper polishing: added examples for probabilistic behavior types

  Adding mutation changes the dynamics and its rest points. Let's assume that $Q_{ji} = .9$
  when $j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
  plotted as the linear green line in Figure~\ref{fig:Updates_RMS}. It has only one stable rest
  point at $x = 0.5$ (see Figure~\ref{fig:Phase_RD}). If we first take the replicator step and
  then the mutation step in sequence, we obtain the replicator mutator dynamic
  $RMD(x) = M(RD(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
  Figure~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
  former two are attractors (see Figure~\ref{fig:Phase_RD}).
\end{example}

@@

While this was only an abstract and simple example, our goal is to apply the RMD to a more
complex case of co-evolving of lexical concepts and pragmatic behavior. To do so, we need to
fix three things: (i) what the relevant types are, (ii) how fitness derives from communicative
success and (iii) how the mutation matrix is computed. These issues are addressed, one by one,
in the following.

\subsection{Types: Lexica and linguistic behavior}
\label{sec:languages+use}

Types are what evolution operates on. In standard applications of evolutionary game theory,
types are ways of acting in a game, e.g., either cooperating or defecting in a prisoner's
dilemma, or, in our previous example, they are what defines an agent's fitness in interaction
with others, such as being positive or negative (whatever that means). For our present
purposes, types are identified more concretely by their cognitive make-up. In general terms, we
are interested in the evolutionary competition between different lexical concepts and ways of
using them in communication. A type in our setting is therefore a pair consisting of a lexicon
and a pragmatic strategy.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a set of
states. A pragmatic type maps a lexicon onto a probabilistic speaker rule (a probabilistic
choice of message for each state) and a probabilistic listener rule (a probabilistic choice of
state for each message). There are many ways of making these general notions more
concrete. Here is what we will assume in the remainder of this paper.

\bigskip
\bigskip
\bigskip

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}.  
Lexica for this case would assign a Boolean truth value, either $0$ for false or $1$ for true, to each state-message pair. The following two lexica exemplify the distinction
between a lexicalized upper-bound for \emph{some} in $\Lbound$ and the widely assumed logical
semantics with only a lower-bound in $\Llack$.

@@

We distinguish between two kinds of linguistic behavior. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
act (approximately) rational with respect to level-$n$ behavior of their
interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
states?} \tb{I think that's fine. $S_n(\cdot \mid \cdot)$ is not used beyond this section. Otherwise we can, e.g., relabel $M$ as $F$(orms) and use $M$(eanings) instead of $S$(tates). I'm fine with either option.}
\begin{flalign}
&H_{0}(s \mid m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m \mid s;L) \propto \exp(\lambda \; L_{sm}) \label{s:level0}
\end{flalign}

According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform. A literal interpreter with lexicon $\Lbound$
from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after
hearing $\msome$, while a literal interpreter with lexicon $\Llack$ would assign $\ssome$
probability $H_0(\ssome \mid \msome; \Llack) = 0.5$. As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
increases, choices made in production are more rational in that they are increasingly in line with expected utility maximization. Expected utility of a message
$\messg$ in state $\state$ for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct meaning. Put differently, expected utility indicates how likely the hearer is thought to infer $\state$ when sent $\messg$. The formulation
in (\ref{s:level0}) also allows false messages to be sent with a small positive probability in
analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is also necessary to guarantee a mutation matrix with only positive entries (see below). If $\lambda = 1$ a literal
speaker with either lexicon $\Lbound$ or $\Llack$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx .27$ for $\Lbound$ and $S_{0}(\msome \mid
\sall;\Llack) = .5$ for $\Llack$. By contrast, if $\lambda = 20$, a literal speaker with either lexicon produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx 1$. But the probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) = .5$ for $\Llack$ because it semantically associates $\msome$ with both states.

Pragmatic behavior of level $n+1$ is similar to its literal counterparts but bases linguistic choice in 
interpretation or production on the expected behavior of a level-$n$ interlocutor instead of on lexical meaning:
\begin{flalign}
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{flalign}

Particularly important to our purpose is the fact non-literal interlocutors using $\Llack$ can {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$,  in contrast to their literal counterparts. Intuitively, this is so because hearers reason a speaker to convey $\sall$ with $\mall$, as this message has a higher chance of leading to mutual understanding than using underspecified $\msome$. Therefore they will expect $\msome$ to be more strongly associated with $\ssome$ than with $\sall$ as it is already possible to clearly convey the latter state $\mall$. The converse reasoning pattern holds for the speaker. In this way, two distinct lexical conventions such as $\Lbound$ and $\Llack$ can give rise to similar observable linguistic behavior. In this case, associating $\msome$ with $\ssome$ and $\mall$ with $\sall$.

\subsection{Fitness \& fitness-based selection based on expressivity}\label{sec:expressivity}

Most evolutionary dynamics assume that the proportion of type $i$ in a population will increase
or decrease as a function of its relative fitness $f_i$. In the context of language evolution,
fitness is frequently associated with expressivity, i.e., the ability to successfully
communicate with other language users from the same population
\citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under a biological
interpretation the assumption is that organisms have a higher chance of survival and
reproduction if they are able to share and receive useful information via communication with
peers. Under a cultural interpretation the picture is that agents themselves strive towards
communicative success and therefore occasionally adapt or revise their behavior to achieve a
higher communicative success (see \citealt[\S3.3]{benz+etal:2005b} for discussion).

The replicator equation gives us the means to make the ensuing dynamics precise, without
necessarily committing to a biological or cultural interpretation. As above, the proportion of types in a
given population is codified in a vector $\vec{x}$, where $x_i$ is type $i$'s proportion. The
fitness of type $i$ is its average expected communicative success, or \emph{expected
  utility} (EU), given the frequencies of types in the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
The expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. Assuming that agents are speakers half
of the time, this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, \text{EU}_S(t_i,t_j) + \nicefrac{1}{2} \, \text{EU}_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
are type $i$'s and type $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} R_{n_j}(s' \mid m;L_j)
  \delta(s,s') \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)
\end{flalign*}
As usual for cooperative communication, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.

\subsection{Learnability}
\label{sec:learnability}

Languages are shaped not only by functionalist forces towards greater expressivity. Another
important factor is the fidelity by which language is transmitted. Among others, linguistic
production can be prone to errors, states or messages may be perceived incorrectly, and
multiple languages may be compatible with the data learners are exposed to. These sources of
uncertainty introduce variation in the transmission of linguistic knowledge from one generation to the next. In particular, learning biases in the iterated transmission process can influence language evolution
substantially.

In biological evolution, where types are expressed genetically, transmission infidelity comes
into the picture through infrequent and mostly random genetic mutations. However, an agent's lexicon
and pragmatic reasoning behavior is not inherited genetically. They need to be learned from
observation. Concretely, when agents of type $j$ want to adopt or imitate the linguistic
behavior of type $i$, they observe the overt linguistic behavior of type $i$ and need to infer what
their type is from that. Iterated learning is a process in which languages are learned repeatedly from the observation
of linguistic behavior of agents who have acquired the language from observation and inference
before as well. In the simplest case there is a single teacher and a single
learner. After sufficient training the learner becomes a teacher and produces behavior that
serves as input for a new learner. Due to the pressure towards learnability it exerts, iterated
learning alone generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014}
and \citealt{tamariz+kirby:2016} for recent surveys).

Following \citet{griffiths+kalish:2007} we model language acquisition as a process of Bayesian
inference in which learners combine the likelihood of a type producing the witnessed learning
input with prior inductive biases. Experimental and mathematical results on iterated learning
suggest that the outcome of this process reflects learners' inductive biases
\citep[e.g.,][]{kirby+etal:2014}. In a Bayesian setting these biases can be codified in a prior
$P \in \Delta(T)$, which reflects the amount of data a learner requires to faithfully acquire
the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}. The extent of the prior's
influence has been shown to heavily depend on the learning strategy assumed to underlie the
inference process. On the one hand, early simulation results suggested that weak biases could
be magnified by exposing learners to only small data samples \citep[e.g. in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
characterization showed that iterated learning converged to the prior, i.e., that the resulting
distribution over languages corresponds to the learners' prior distribution and is not
influenced by the amount of input given to them. This difference in predictions can be traced
back to differences in the selection of hypotheses from the posterior. Griffith \& Kalish's
convergence to the prior holds for learners that sample from the posterior. More deterministic
strategies such as the adoption of the type with the highest posterior probability, so-called
{\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the
data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a parameter
$l\ge1$ to modulate between posterior sampling and the MAP strategy. When $l = 1$ learners
sample from the posterior. The learners' propensity to maximize the posterior grows as $l$
increases.

Let $D$ be the set of possible data that learners may be exposed to. This set $D$ contains all
sequences of state-message pairs of length $k$, e.g.,
$\tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$. As $k$ increases, learners have more data to base their inference on and so tend to
recover the true types that generated a given sequence with higher probability. The mutation
matrix $Q$ of the replicator mutator dynamics in (\ref{eq:RMD_discrete}) can then be defined as
follows: $Q_{ji}$ is the probability that a learner acquires type $i$ when learning from an
agent of type $j$. The learner observes a length-$k$ sequence $d$ of state-message pairs, but
the probability $P(d \mid t_j)$ with which sequence $d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$ is observed depends on type $j$'s
behavior:
\begin{align*}
  P(d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}} \mid t_j) = \prod_{i = 1}^k S_{n_j}(m_i
  \mid s_i; L_{j})\,,
\end{align*}
where, as before, $n_j$ is $j$'s pragmatic reasoning type and $L_j$ is $j$'s lexicon. For a
given observation $d$, the probability of acquiring type $i$ is $F(t_i \mid d)$, so that:
\begin{flalign*}
  Q_{ji} \propto \sum_{d \in D} P(d \mid t_j) F(t_i \mid d)\,.
\end{flalign*}
The acquisition probability $F(t_i \mid
d)$ given datum $d$ is obtained by probability matching $l = 1$ or a tendency towards choosing
the most likely type $l > 1$ from the posterior distribution $P(\cdot \mid d)$ over types given
the data, which is calculated by Bayes' rule:
\begin{flalign*}
  & F(t_i \mid d) \propto P(t_i \mid d)^l \; \text{ and }\\
  & P(t_i \mid d) \propto P(t_i) P(d \mid t_i)\,.
\end{flalign*}

>>>>

  Adding mutation changes the dynamic and its rest points. Let's assume that $Q_{ji} = .9$
  when $j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
  plotted as the linear green line in Figure~\ref{fig:Updates_RMS}. It has only one stable rest
  point at $x = 0.5$ (see Figure~\ref{fig:Phase_RD}). If we first take the replicator step and
  then the mutation step in sequence, we obtain the replicator mutator dynamic
  $RMD(x) = M(RD(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
  Figure~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
  former two are attractors (see Figure~\ref{fig:Phase_RD}).
\end{example}

@@

\subsection{Fitness \& learnability of lexical meanings \& pragmatic strategies}
\label{sec:fitn--learn}

Moving beyond an abstract example, our goal is to apply the RMD to investigate the co-evolution
of lexical concepts and pragmatic behavior. To do so, we need to fix three things: (i) what the
relevant types are, (ii) how fitness derives from communicative success and (iii) how the
mutation matrix is computed. These issues are addressed, one by one, in the following.

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}

Types are what evolution operates on. They define an agent's fitness, usually through a payoff
accrued in single interactions with other agents. Often types can be identified as the possible
acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
they may be thought of as general properties of an agent that influences her fitness, such as
being positive or negative in our previous example (whatever that means). For our present
purposes, types are identified more concretely by specific assumptions about their cognitive
make-up. Since we are interested in the evolutionary competition between different lexical
concepts and ways of using them in communication, a type is here defined as a pair consisting
of a lexicon and a pragmatic strategy.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
(possibly fuzzy) set of states. A pragmatic type maps a lexicon onto a probabilistic speaker
rule (a probabilistic choice of message for each state) and a probabilistic listener rule (a
probabilistic choice of state for each message). There are many ways of making these general
notions more concrete. Here is what we will assume in the remainder of this paper.

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}.  Lexica for this case would assign a
Boolean truth value, either $0$ for false or $1$ for true, to each state-message pair. The
following two lexica are minimal examples for the distinction between a lexicalized upper-bound
for \emph{some} in $\Lbound$ and the widely assumed logical semantics with only a lower-bound
in $\Llack$.

@@

Pragmatic types define dispositions of produce and interpret messages given a lexicon. We
distinguish between two kinds of pragmatic types. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use, as in Equations~(\ref{h:level0}) and
(\ref{s:level0}). Pragmatic language users of level $n + 1$ act (approximately) rational with
respect to level-$n$ behavior of their interlocutors, as in Equations~(\ref{h:leveln}) and
(\ref{s:leveln}).
\begin{alignat}{3}
&H_{0}(s \mid m;L) &&\propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m \mid s;L) &&\propto \exp(\lambda \; L_{sm}) \label{s:level0}\\
&H_{n+1}(s|m;L) &&\propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) &&\propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{alignat}

According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat, for
simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
message. Pragmatic hearers, defined in (\ref{h:leveln}), use Bayes rule to weigh
interpretations based on a conjecture about speaker behavior. Speaker behavior is regulated by
a soft-max parameter $\lambda$, $\lambda \geq 0$ \citep{luce:1959,sutton+barto:1998}. As
$\lambda$ increases, choices approximate strict maximization of expected utilities. Expected
utility of a message $\messg$ in state $\state$ for a level $n+1$ speaker is here defined as
$H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct
meaning. For literal speakers, utility only tracks truthfulness. Literal speakers choose any
true message with equal probability but may send false messages as well with a probability
dependent on $\lambda$.

Here are some concrete examples. \mf{please double-check the numbers!} A literal interpreter with lexicon $\Lbound$ assigns $\ssome$
a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after hearing $\msome$, while a literal
interpreter with $\Llack$ has $H_0(\ssome \mid \msome; \Llack) = 0.5$:
\begin{align*}
  H_0(\cdot \mid \cdot, \Lbound) & = \begin{blockarray}{lcc}
    & \mygray{\ssome} & \mygray{\sall} \\
    \begin{block}{l[cc]}
      \mygray{\msome} & 1 & 0 \\
      \mygray{\mall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  H_0(\cdot \mid \cdot, \Llack) & = \begin{blockarray}{lcc}
    & \mygray{\ssome} & \mygray{\sall} \\
    \begin{block}{l[cc]}
      \mygray{\msome} & .5 & .5 \\
      \mygray{\mall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} 
\end{align*}
If $\lambda = 1$ literal speaker behavior looks like this:
\begin{align*}
  S_0(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & .73 & .27 \\
      \mygray{\sall}  & .27 & .73 \\
    \end{block}
  \end{blockarray} &
  S_0(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & .73 & .27 \\
      \mygray{\sall}  & .5 & .5 \\
    \end{block}
  \end{blockarray} 
\end{align*}
If $\lambda = 20$, literal speakers will choose with less slack:
\begin{align*}
  S_0(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  S_0(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} &  1 & 0 \\
      \mygray{\sall}  & .5 & .5 \\
    \end{block}
  \end{blockarray} 
\end{align*}
Pragmatic hearers of level 1 have the following probabilistic interpretation behavior for
$\lambda = 1$:
\begin{align*}
  H_1(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
    & \mygray{\ssome} & \mygray{\sall} \\
    \begin{block}{l[cc]}
      \mygray{\msome} & .73 & .27 \\
      \mygray{\mall}  & .27 & .73 \\
    \end{block}
  \end{blockarray} &
  H_1(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
    & \mygray{\ssome} & \mygray{\sall} \\
    \begin{block}{l[cc]}
      \mygray{\msome} & .59 & .41 \\
      \mygray{\mall}  & 0.35 & 0.65 \\
    \end{block}
  \end{blockarray} 
\end{align*}
For $\lambda = 20$, there wil less slack again:
Pragmatic listeners of level 1 have the following probabilistic interpretation behavior for
$\lambda = 1$:
\begin{align*}
  H_1(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
    & \mygray{\ssome} & \mygray{\sall} \\
    \begin{block}{l[cc]}
      \mygray{\msome} & 1 & 0 \\
      \mygray{\mall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  H_1(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
    & \mygray{\ssome} & \mygray{\sall} \\
    \begin{block}{l[cc]}
      \mygray{\msome} & 0.67 & 0.33 \\
      \mygray{\mall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} 
\end{align*}
With $\lambda = 1$, pragmatic speakers of level 1 will have:
\begin{align*}
  S_1(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & .73 & .37 \\
      \mygray{\sall}  & .37 & .73 \\
    \end{block}
  \end{blockarray} &
  S_1(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & .62 & .38 \\
      \mygray{\sall}  & .38 & .62 \\
    \end{block}
  \end{blockarray} 
\end{align*}
For high $\lambda = 20$, pragmatic speakers of level 1 give us:
\begin{align*}
  S_1(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  S_1(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} 
\end{align*}

It is important to note that pragmatic agents of level 1 using $\Llack$ associate $\msome$
preferentially with $\ssome$, in contrast to their literal counterparts of level 0. This
association is not perfect, and usually less strong than what agents with a lexicalized upper
bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher oder iteration of
reasoning types leads to stronger associations of $\msome$ and $\ssome$ also for the
hearer. Still, the case study presented in Section~\ref{sec:si-case-study} will consider
speaker and hearer behavior at levels 0 and 1, as the latter are the simplest pragmatic
reasoning types which show a tendency to communicatively attuned pragmatic enrichment. Using
only level 1 reasoning types is therefore a conservative choice that works against the
fitness-based selection of pragmatic language use for a notion of fitness based on
communicative efficiency, which is introduced next.

\subsubsection{Fitness \& fitness-based selection based on expressivity}\label{sec:expressivity}

Under the replicator dynamic the proportion of type $i$ in a population will increase or
decrease as a function of its relative fitness $f_i$. In the context of language evolution,
fitness is usually associated with expressivity, i.e., the ability to successfully communicate
with other language users from the same population
\citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under a biological
interpretation the assumption is that organisms have a higher chance of survival and
reproduction if they are able to share and receive useful information via communication with
peers. Under a cultural interpretation the picture is that agents themselves strive towards
communicative success and therefore occasionally adapt or revise their behavior to achieve
higher communicative success (see \citealt[\S3.3]{benz+etal:2005b} for discussion).

The replicator equation gives us the means to make the ensuing dynamic precise, without
necessarily committing to a biological or cultural interpretation. As above, the proportion of
types in a given population is codified in a vector $\vec{x}$, where $x_i$ is the proportion of
type $i$. The fitness of type $i$ is its average expected communicative success, or
\emph{expected utility} (EU), given the frequencies of types in the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
The expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. If agents are speakers half of the
time, this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, \text{EU}_S(t_i,t_j) + \nicefrac{1}{2} \, \text{EU}_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
are type $i$'s and type $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} H_{n_j}(s' \mid m;L_j)
  \delta(s,s') \,, \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)\,.
\end{flalign*}
As usual, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.

\subsubsection{Learnability}
\label{sec:learnability}

Languages are shaped not only by functionalist forces towards greater expressivity. Another
important factor is the fidelity with which linguistic knowledge is transmitted. Among others,
linguistic production can be prone to errors, states or messages may be perceived incorrectly,
and multiple languages may be compatible with the data learners are exposed to. These sources
of uncertainty introduce variation in the transmission of linguistic knowledge from one
generation to the next. In particular, learning biases in the iterated transmission process can
influence language evolution substantially.

In biological evolution, where types are expressed genetically, transmission infidelity comes
into the picture through infrequent and mostly random genetic mutations. However, an agent's
lexicon and pragmatic reasoning behavior is likely not inherited genetically. They need to be
learned from observation. Concretely, when agents of type $j$ want to adopt or imitate the
linguistic behavior of type $i$, they observe the overt linguistic behavior of type $i$ and
need to infer the covert type that most likely produced the visible behavior. Iterated learning
is a process in which languages are learned repeatedly from the observation of linguistic
behavior of agents who have themselves acquired the language from observation and inference. In
the simplest case there is a single teacher and a single learner. After sufficient training the
learner becomes a teacher and produces behavior that serves as input for a new learner. Due to
the pressure towards learnability it exerts, iterated learning alone generally leads to simpler
and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for
recent surveys).

Following \citet{griffiths+kalish:2007} we model language acquisition as a process of Bayesian
inference in which learners combine the likelihood of a type producing the witnessed learning
input with prior inductive biases. Experimental and mathematical results on iterated learning
suggest that the outcome of this process reflects learners' inductive biases
\citep[e.g.,][]{kirby+etal:2014}. In a Bayesian setting these biases can be codified in a prior
$P \in \Delta(T)$, which reflects the amount of data a learner requires to faithfully acquire
the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}. The extent of the prior's
influence has been shown to heavily depend on the learning strategy assumed to underlie the
inference process. On the one hand, early simulation results suggested that weak biases could
be magnified by exposing learners to only small data samples
\citep[e.g. in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
characterization showed that iterated learning converges to the prior in the limit, i.e., that
the resulting distribution over languages corresponds to the learners' prior distribution and
is not influenced by the amount of input given to them. This difference in predictions can be
traced back to differences in the selection of hypotheses from the posterior. Griffith \&
Kalish's convergence to the prior holds for learners that sample from the posterior. More
deterministic strategies such as the adoption of the type with the highest posterior
probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of
both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we
use a parameter $l\ge1$ to modulate between posterior sampling and the MAP strategy. When
$l = 1$ learners sample from the posterior. The learners' propensity to maximize the posterior
grows as $l$ increases.

Let $D$ be the set of possible data that learners may be exposed to. This set $D$ contains all
sequences of state-message pairs of length $k$, e.g.,
$\tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$. As $k$ increases, learners have more data to base their inference on and so tend to
recover the true types that generated a given sequence with higher probability. The mutation
matrix $Q$ of the replicator mutator dynamic in (\ref{eq:RMD_discrete}) can then be defined as
follows: $Q_{ji}$ is the probability that a learner acquires type $i$ when learning from an
agent of type $j$. The learner observes a length-$k$ sequence $d$ of state-message pairs, but
the probability $P(d \mid t_j)$ with which sequence $d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$ is observed depends on type $j$'s
behavior:
\begin{align*}
  P(d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}} \mid t_j) = \prod_{i = 1}^k S_{n_j}(m_i
  \mid s_i; L_{j})\,,
\end{align*}
where, as before, $n_j$ is $j$'s pragmatic reasoning type and $L_j$ is $j$'s lexicon. For a
given observation $d$, the probability of acquiring type $i$ is $F(t_i \mid d)$, so that:
\begin{flalign*}
  Q_{ji} \propto \sum_{d \in D} P(d \mid t_j) F(t_i \mid d)\,.
\end{flalign*}
The acquisition probability $F(t_i \mid
d)$ given datum $d$ is obtained by probability matching $l = 1$ or a tendency towards choosing
the most likely type $l > 1$ from the posterior distribution $P(\cdot \mid d)$ over types given
the data, which is calculated by Bayes' rule:
\begin{flalign*}
  & F(t_i \mid d) \propto P(t_i \mid d)^l \; \text{ and }\\
  & P(t_i \mid d) \propto P(t_i) P(d \mid t_i)\,.
\end{flalign*}
